Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-22 paper/ms.tex
more small edits and comments in the intro

% TO DOS
\newcommand{\todo}[3]{{\color{#2}\emph{#1}: #3}}
\newcommand{\dfmtodo}[1]{\todo{DFM}{red}{#1}}
\newcommand{\agoltodo}[1]{\todo{Agol}{blue}{#1}}

@@

In the following Section, we present a method for substantially improving this
scaling in many circumstances.
We call our method and its implementations \celerite.\footnote{The name
\celerite\ comes from the French word \foreign{c\'elerit\'e} meaning the speed
of light in a vacuum.} The \celerite\ method requires using a specific model
for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and, although it
has some limitations, we demonstrate in subsequent sections that it can
increase the computational efficiency of many astronomical data analysis
problems.
The main limitation of this method is that it can only be applied to
one-dimensional datasets, where by ``one-dimensional'' we mean that the
 \emph{input coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv
t_n$.\footnote{We are using $t$ as the input coordinate because
one-dimensional GPs are often applied to time series data but this isn't a
real restriction and the \celerite\ method can be applied to \emph{any}
one-dimensional dataset.}
Furthermore, the covariance function for the \celerite\ method is
``stationary'', \ie\ $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function of
$\tau_{nm} \equiv |t_n - t_m|$.

@@

This kernel function can be made even more general by
introducing complex parameters $a_j \to a_j\pm i\,b_j$ and
$c_j \to c_j\pm i\,d_j$.
In this case, the covariance function becomes
\begin{eqnarray}\eqlabel{celerite-kernel-complex}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau_{nm}\right)
        \right. \nonumber\\
    &&+\left.
    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau_{nm}\right)
\right]
\end{eqnarray}
and, for this function, \eq{gp-likelihood} can still be evaluated with
$\mathcal{O}(N)$ operations.
The details of this method and a few implementation considerations are
discussed in the following \sectionname\ but we first discuss some
properties of this covariance function.

@@

Finally, we note that the product of two terms of the form found inside the
sum in \eq{celerite-kernel} can also be re-written as a sum with updated
parameters
\begin{eqnarray}\eqlabel{product-rule}
k_j(\tau) \, k_k(\tau) =
    e^{-\tilde{c}\,\tau}\,[
        \tilde{a}_+\,\cos(\tilde{d}_+\,\tau) + \tilde{b}_+\,\sin(\tilde{d}_+\,\tau) +
        \tilde{a}_-\,\cos(\tilde{d}_-\,\tau) + \tilde{b}_-\,\sin(\tilde{d}_-\,\tau)
    ]
\end{eqnarray}
where
\begin{eqnarray}
    \tilde{a}_{\pm} &=& \frac{1}{2}\,(a_j\,a_k \pm b_j\,b_k) \\
    \tilde{b}_{\pm} &=& \frac{1}{2}\,(b_j\,a_k \mp a_j\,b_k) \\
    \tilde{c} &=& c_j + c_k \\
    \tilde{d}_{\pm} &=& d_j \mp d_k \quad.
\end{eqnarray}
Therefore, the method described in the following section can be used to scale
inferences to large datasets for any model where the kernel function is a
general sum or product of \celerite\ terms.

>>>>

% TO DOS
\newcommand{\todo}[3]{{\color{#2}\emph{#1}: #3}}
\newcommand{\dfmtodo}[1]{\todo{DFM}{red}{#1}}
\newcommand{\agoltodo}[1]{\todo{Agol}{blue}{#1}}
\newcommand{\racomment}[1]{{\color{magenta}#1}}

@@

In the following Section, we present a method for substantially improving this
scaling in many circumstances.
We call our method and its implementations \celerite.\footnote{The name
\celerite\ comes from the French word \foreign{c\'elerit\'e} meaning the speed
of light in a vacuum.} The \celerite\ method requires using a specific model
for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and, although it
has some limitations, we demonstrate in subsequent sections that it can
increase the computational efficiency of many astronomical data analysis
problems.
The main limitation of this method is that it can only be applied to
one-dimensional datasets, where by ``one-dimensional'' we mean that the
 \emph{input coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv
t_n$.\footnote{We use $t$ as the input coordinate because
one-dimensional GPs are often applied to time series data but this isn't a
real restriction and the \celerite\ method can be applied to \emph{any}
one-dimensional dataset.}
Furthermore, the covariance function for the \celerite\ method is
``stationary'', \ie\ $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function of
$\tau_{nm} \equiv |t_n - t_m|$.

@@

This kernel function can be made even more general by
introducing complex parameters $a_j \to a_j\pm i\,b_j$ and
$c_j \to c_j\pm i\,d_j$.
In this case, the covariance function becomes
\begin{eqnarray}\eqlabel{celerite-kernel-complex}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau_{nm}\right)
        \right. \nonumber\\
    &&+\left.
    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau_{nm}\right)
\right]
\end{eqnarray}
and, for this function, \eq{gp-likelihood} can still be evaluated with
$\mathcal{O}(N)$ operations. \racomment{Not $\mathcal{O}(NJ^2)$?}
The details of this method and a few implementation considerations are
outlined in the following \sectionname\ but we first discuss some
properties of this covariance function.

@@

Finally, we note that the product of two terms of the form found inside the
sum in \eq{celerite-kernel} can also be re-written as a sum with updated
parameters
\begin{eqnarray}\eqlabel{product-rule}
k_j(\tau) \, k_k(\tau) =
    e^{-\tilde{c}\,\tau}\,[
        \tilde{a}_+\,\cos(\tilde{d}_+\,\tau) + \tilde{b}_+\,\sin(\tilde{d}_+\,\tau) +
        \tilde{a}_-\,\cos(\tilde{d}_-\,\tau) + \tilde{b}_-\,\sin(\tilde{d}_-\,\tau)
    ]
\end{eqnarray}
where
\begin{eqnarray}
    \tilde{a}_{\pm} &=& \frac{1}{2}\,(a_j\,a_k \pm b_j\,b_k) \\
    \tilde{b}_{\pm} &=& \frac{1}{2}\,(b_j\,a_k \mp a_j\,b_k) \\
    \tilde{c} &=& c_j + c_k \\
    \tilde{d}_{\pm} &=& d_j \mp d_k \quad.
\end{eqnarray}
Therefore, the method described in the following section can be used to
perform scalable inference on large datasets for any model where the kernel
function is a sum or product of \celerite\ terms.
